ppseq: An R Package for Sequential Predictive Probability Monitoring

### 2022-01-13 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
paper edits and some man fixes

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib

@@

Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error. As a result, predictive probability has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014; @Hobbs2018]. However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design. This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring. Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} compare designs based on different thresholds for decision making. Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability design using the \pkg{ppseq} package. While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential clinical trial design.

@@

The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution. Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$. Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$. Using the posterior probability, which represents the probability of success based only on the data accrued so far, we would declare a treatment efficacious if $\Pr(p>p_0 | X) > \theta$, where $\theta$ represents a pre-specified posterior decision threshold. The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$. Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$. The posterior predictive probability represents the probability that the treatment will be declared efficacious at the end of the trial when full enrollment is reached. We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e. $PPP<\theta^*$. Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success. Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.

@@

The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds. Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}. Default arguments are available for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}, but customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

@@

To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function. The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, and the predictive threshold of the selected design \code{ppp}. Default arguments are provided for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}, though customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case). The function results in a \code{tibble}. In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}. The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$. At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$. In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively. There are also columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively. Finally, there is a column for the posterior predictive probability associated with that decision \code{ppp}.

@@

In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796. This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment. Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial. Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package. In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial. After the first 10 patients we would stop the trial if there were 0 responses, and so on. At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were >=14 responses.

>>>>

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib

@@

Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error. As a result, predictive probability has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014; @Hobbs2018]. However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design. This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility. Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making. Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \pkg{ppseq} package. While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.

@@

The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution. Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$. Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$. A posterior probability threshold $\theta$ would be pre-specified during the trial design stage. At the end of the trial if the posterior probability exceeded the pre-specified threshold, i.e. if $\Pr(p>p_0 | X) > \theta$, the trial would be declared a success.

The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$. Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$. The posterior predictive probability represents the probability that, at any given interim monitoring point, the treatment will be declared efficacious at the end of the trial when full enrollment is reached, conditional on the currently observed data and the specified priors. We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e. $PPP<\theta^*$. Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success. Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.

@@

The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility. The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds. Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}. Default arguments are available for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}, but customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

@@

To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function. The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, and the predictive threshold of the selected design \code{ppp}. Default arguments are provided for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}, though customization is also possible. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case). The function results in a \code{tibble}. In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}. The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$. At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$. In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively. There are also columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively. Finally, there is a column for the posterior predictive probability associated with that decision \code{ppp}.

@@

In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796. This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment. Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial. Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package. In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial. After the first 10 patients we would stop the trial if there were 0 responses, and so on. At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were greater than or equal to 14 responses.
